Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-01-15 text/front_end_v2.qmd
working through results, homie

---
title: "Structuring College Access: The Market Segment Model and College Board Geo-markets"
author: ["Ozan Jaquette (UCLA)", "Karina Salazar (University of Arizona)"]
bibliography: ./../assets/bib/cb_geomarket_bib.bib
csl: ./../assets/bib/apa.csl
format:
  pdf:
    documentclass: article
    #papersize: letter
    number-sections: true
    number-depth: 2
    #toc: FALSE
    #pandoc_args: !expr rmdfiltr::add_wordcount_filter(rmdfiltr::add_citeproc_filter(args = NULL))
    #keep_tex: true
    extensions: [andrewheiss/wordcount]
editor: visual
geometry:
  margin=1in
fontsize: 11pt
header-includes:
  - \usepackage{setspace}
  - \doublespacing
---

@@

In 1983, Robert Zemsky and Penney Oedel authored *The Structure of College Choice*. Based on an analysis of 1980 SAT score-sending behavior and published by *The College Board*, @RN4982 developed the Market Segment Model to help colleges recruit students. The underlying idea of the Market Segment Model is that student demand for your institution is fundamentally a function of class. Therefore, the key to recruitment is identifying territories where large numbers of your class live. @RN4982[p. 11] describe the Market Segment Model as an effort to "quantify" the knowledge that college admissions officers have about the geography of student demand:

@@

Geomarkets have been incorporated into the supply-side structure of college access in ways that are simultaneusly mundane and fundamental. We posit three ways. First, Geomarkets became a principle for how many -- if -- admissions offices organized their labor. In states where a college recruits heavily from, admissions officers are often assigned specific Geomarkets as their territory [CITE]. Geomarkets were created to mirror the territories of admissions officers and, in turn, admissions offices structured their territories around Geomarkets. Second, The Market Segment Model and Geomarkets became the basis for Geomarkets are the basis for the College Board Enrollment Planning Service (EPS), founded in 1984 and still active today. EPS software recommends which Geomarkets a college should recruit from and which schools/communities they should prioritize within targeted Geomarkets. @RN5019_2010 says "Enrollment Planning Service (EPS\textsuperscript{\textregistered}) the analysis tool that pinpoints the schools and geomarkets where your best prospects are most likely to be found." @RN5021 reports that, in 1995, 37% of 4-year public institutions and 49% of 4-year private institutions used EPS.

@@

Geomarkets have never been the subject of empirical research. This omission is surprising because the sociology of education is concerned with sorting [@RN4873], particularly sorting into college. Traditionally, most scholarship -- and certainly most scholarship in economics [HOXBY? CHETTY] -- views inequality in college access as a function households or k-12 schools, which represent the demand-side of higher education. More recently, scholarship in the sociology of education takes seriously the idea that colleges are not passive recipients of applications, but rather they expend substantial resources soliciting demand from desirable applicants [@RN3519; @RN4774; @RN4324]. However, most scholarship on recruiting analyzes the behavior of individual colleges [e.g., @RN4758; @RN4759], thereby creating the assumption that recruiting is a function of individual colleges. What these studies have missed is the broader enrollment management industry -- third-party vendors, their products and consulting solutions -- as a set of mechanisms that structure college access, funneling certain kinds of students to certain kinds of institutions.

This manuscript analyzes one of those mechanisms, The Market Segment Model and College Board Geomarkets, as a case study of quantification. We develop arguments from the broader literature on quantification \[e.g.,\] [@RN5003; @RN5005; @RN5007], particularly the discussions of correlation and homophily from @RN4975. Predictive analytics are based on correlation. When we use data on past correlations to make recommendations about the future, we amplify the effects of historic structural inequality. The Market Segment Model analyzes SAT score-sending data from 1980 and concludes that student demand for higher education is primarily a function of social class [@RN4982]. The Market Segment Model argues that homophily -- actors that share characteristics are likely to form connections --  is the organizing principle of student demand and competition between colleges. @RN4982[p. 72] state that "the hierarchical structure of collegiate competition largely reflects the stratified social and economic dimensions of the communities from which colleges draw their students."  This snapshot of student demand in 1980, itself a consequence historic structural inequality, was programmed into the way admissions offices organize labor and programmed into the EPS and Student Service Service products that colleges utilize to identify and target prospective students. The result is a supply-side that reinforces structural inequalities observed on the demand-side.

@@

<!-- 
THIS TEXT IS FINE, BUT TRYING TO SET UP THE CONTRIBUTION TO BROADER QUANTIFICATION LITERATURE, VIS-A-VIS NO-ONE HAS STUDIED MARKET RESEARCH; DON'T THINK THIS IS NECESSARY AND YOU ARE NOT ON FIRM GROUND HERE IN CALLING THIS "MARKET RESEARCH"
Scholarship on quantification in education has focused on consumer-facing measures such as U.S. News Rankings [@RN5003] and UK "league tables" about school quality [@RN5005]. To our knowledge, prior knowledge has not examined business-facing measures produced by third-party market research. Here, the desirable attributes of students and places they live are quantified with the purpose of informing colleges about where they should allocate recruiting resources to efficiently enroll the most desirable students. We reason that market research reproduces inequality by encouraging colleges to plan their recruiting efforts based on a snapshot of social stratification in student demand that is substantially a function of historic, structural inequality in educational opportunity. In contrast to @RN5005, we cannot demonstrate the effect of geomarkets on inequality. However, we show the mechanisms of how geomarkets reproduce inequality by simulating who is included/excluded when geomarkets are utilized within student list purchases as recommended by @RN4982.
-->

@@

@@

Scholarship at the nexus of enrollment management and college access can be categorized by which part(s) of the enrollment funnel it speaks to. The majority of scholarship focuses on the admissions stage, analyzing which admissions criteria are utilized and/or which applicants are admitted [e.g., @RN4874; @RN4993; @RN4131; @RN3522; @poon2022rethinking]. Economists often investigate financial aid leveraging, which seeks to convert admits to enrolled students [CITE]. 
<!--  Scholarship on college access often takes a demand-side perspective, whereby household differences in economic and cultural capital yield stratification in student preferences and the ability to gain admission [e.g., @RN4935; @RN1814].   -->

@@

Scholarship about U.S. News & World Report (USNWR) Law School rankings demonstrate the reactivity, discipline, and authority effects of quantification [@RN5003; @RN2356; @RN2376; @RN2375]. Law school rankings were developed as a means of informing prospective students about the relative quality of different programs. Speaking to *reactivity* effects, rankings had the intended effect of affecting where students apply and enroll [CITE]. However, rankings also affected hiring decisions of firms because clients evaluated firms based on the prestige of law schools attended by firm lawyers. Once law schools realized that "important groups of constituents --- students, faculty, trustees, employers, other media --- were using rankings to make decisions that had large consequences for schools \ldots \[then\] schools felt pressured to take them seriously" [@RN5007, p. 415]. Rankings *disciplined* the behavior of law schools. For example, applicants with characteristics valued by the rankings system (e.g., LSAT scores) became more important for decisions about admissions and merit aid [@RN5003]. Rankings also weakened the *authority* of admissions directors and admissions officers to make independent decisions about which applicants to admit, ideal class size, and curricular offerinags [@RN5003]. <!--
In a process akin to sports analytics in Moneyball \[CITE MICHAEL LEWIS\], we suggest that market research replaces the wisdom and authority of local admissions counselors.    -->

@@

<!-- THIS IS THE CUT TRANSITION PARAGRAPH
In the broader quantification literature, critical data studies [e.g., @RN4975; @RN4849; @RN4772] and sociological analyses of technology and inequality [e.g., @RN4778; @RN4786; @RN4877; @RN4810] introduce concepts salient to the analysis of market research. Our focus is on "selection devices," which utilize some combination of inputs and discretion to categorize places, people, and organizations for the purpose of allocating opportunities or interventions [@RN4778; @RN5012].
-->

@@

Several studies show that using correlational models make predictions reproduces structural inequality [@RN4794]. The correlations observed during the training data stage are a snapshot of relationships between variables at a particular point of time. These correlations may be a function of enduring structural inequality, but underlying causes are not considered by applications of predictive models. Reviewing scholarship about algorithms, @RN4794 [p. 224] state that "predicting the future on the basis of the past threatens to reify and reproduce existing inequalities." Disproportionately targeted/excluded populations are predicted to have a higher risk of an outcome, which amplifies subsequent targeting/exclusion. This phenomenon whereby has been termed the "ratchet effect" [@RN4851] and "pernicious feedback loops" [@RN4849].^[An often cited example is the LSI-R recidivism model which predicts a prisoner's chances of re-arrest and is used by 24 states [@RN4849]. Because the algorithm uses zip code as an input, people who live in highly policed neighborhoods have a higher predicted probability of being arrested, which leads to more policing in those neighborhoods, which perpetuates racialized inequality in arrests. Note that predictive models such as the LSI-R model are not merely used for social science purposes. Rather, they reproduce structural inequality because they direct the allocation of future resources based on a snapshot of historical inequality.]

**Homophily**. @RN4975 argues that homophily is central to models that use correlation to predict future behavior. Homophily means that actors who share common characteristics are likely to form connections with one another, or "birds of a feather flock together." Homophily is a core assumption for network science [@mcpherson2001birds], in which actors -- in this manuscript, high school students -- are linked to one another through direct and indirect network ties -- submitting SAT scores to the same college. Because network science models often draws from rational choice theory, they assume that homophily is the result of voluntary action by individuals, thereby "eras\[ing\] historical contingencies, institutional discrimination, and economic realities" [@RN4975, p. 95] that underlie behavior consistent with homophily. @RN4975 problematizes the idea that homophily is a naturally occurring phenomenon. In commercial social networks (e.g., Faceboo, X, TikTok), homophily is more than an assumption; homophily is programmed into the algorithms that create connections between users.[^2]

[^2]: Thus, @RN4975 [p. 82] writes, "echo chambers are not unfortunate errors, but deliberate goals" because "homophily is used to create agitated clusters of individuals whose angry similarity and overwhelming attrication to their common object of hatred both repel them from one another and glue them together."

Within the market research profession, homophily is central to geodemography and the creation of market segments. Geodemography emerged in the 1970s as a branch of market research that estimates the behavior of consumers based on where they live [@RN4878]. Market segments are subgroups within a larger market that have similar consumer demand. Early geodemographic classifications of consumers (e.g., PRIZM by Claritas Corporation) were derived from publicly available Census data, which disaggregated data to the zip code level. The Claritas Corporation had a financial incentive to argue that people living near one another share similar consumer preferences because geographic localities could then be categorized into market segments that would be useful for direct mail marketing campaigns [@RN4957]. Later, the development of individual credit scores (e.g., FICO score) enabled merchants to classify consumers into many, fine-grained groups [@RN4877]. @RN4810 introduce the concept "classification situations" to describe the expansion of actuarial techniques to categorize customers into many, ordinally ranked groups. Merchants and lenders began tying these classifications to tiered products that targets different consumer groups with different levels of benefits and costs [@RN5015].^[For example, "payday loans" charge high interest rates to consumer groups that were previously denied credit altogether.] Classification situations engender markets where a vertical hierarchy of products are targeted to a vertical hierarchy of consumers. <!--
ORIGINAL TEXT ABOUT GEODEMOGRAPHIC CLASSIFICATION SYSTEMS

@@

Reviewing scholarship on quantification, @RN5006 argue that quantification has effects to the extent that stakeholders care about the numbers. U.S. law school rankings [@RN5003] and UK school league tables [@RN5005] exemplify this sort of salience. Studies of quantification in the sociology of education tend to study consumer-facing metrics. Law school rankings can be conceived as a business-to-consumer product because the goal is to inform consumers (prospective law school students) and consumers pay for the product. By contrast, market research helps businesses identify/target customers and can be conceived as a business-to-business (B2B) product because businesses pay for market research. To our knowledge, scholarship on quantification in education has not examined how third-party market research classifies consumers. Additionally, scholarship on college access has failed to analyze third-party products that colleges use to identify and target prospective students. This manuscript analyzes Market Segment Model [@RN4982], which categorized high school students into vertical market segments and simultaneously created local Geomarkets that could be evaluated based on their composition of student market segments. The Market Segment Model and Geomarkets became the basis for the College Board *Enrollment Planning Service* (EPS), which advised colleges which geo-markets to target. Later, Geomarkets were incorporated into the College Board student list product named Student Search Service.

@@

<!--
The market segment model became the basis for the College Board *Enrollment Planning Service* (EPS), which advised colleges which geo-markets to target. @RN5021 reports that in 1995, 37% of 4-year public universities and 49% of 4-year private universities used EPS, while 41% of 4-year publics and 16% of 4-year privates used ACT's market analysis service product, which was similar to EPS. Geo-markets later became a filter in the College Board *Student Search Service* student list product, which is utilized by the vast majority of public and private universities.

By contrast, many quantification projects -- for example, human rights indicators -- have "little oomph" [@RN5006, p. 265]. FROM BERMAN & HIRSCHAN PAGE 263: "Merry demonstrates how human trafficking indicators largely sit on the shelf because no audience for them has been created." 
-->

# REVIEW OF EEPA 2024 PROLLY GOES IN "THE CASE"

@RN5012 deconstruct the College Board *Student Search Service* product. Selection devices are procedures or routines for making categorical decisions based on input values [@RN4778]. Student list products are selection devices that enable colleges to choose which prospective students they target or ignore based on search filters that incorporate search filters (e.g., high school graduating class, state) which enable colleges to select prospective students. "Racialized inputs" are ostensibly race-neutral inputs that are systematically correlated with race because marginalized racial/ethnic groups have historically been excluded from the input [@RN4786]. @RN5012 argue that several frequently utilized student list filters (zip code, AP test score, SAT score) are racialized inputs. Using a national sample of high school students, and also using data from actual student lists purchased by public universities, they show that racialized search filters have a strong negative relationship with the selection of Black and Latinx prospects. <!-- Using data from actual student lists purchased by public universities, they show that raciaized search filters -- used in conjunction with other search filters -- yield dramatic racial inequality in which prospective students are targeted. -->

Student list products offer many geographic filters. Geographic borders are perhaps the most commonly studied racialized input in scholarship about algorithmic bias [@RN4775; @RN4849; @RN4887], owing to extreme racial segregation in the U.S. In addition to offering geographic filters based on known geographic borders (e.g., zip code, county, CBSA, state), the College Board created new geographic borders that subsequently became filters in the *Student Search Service*. The geodemographic segment filters utilize cluster analysis to allocate each census tract and each high school into different categories based past college enrollment and other factors [@RN4565]. Geodemographic search filters can be read as the mimicry of corporate market logic [@RN4004] in that they apply the logic and methods of commercial geodemographic market segmentation products, such as PRIZM by the Claritas Corporation [@RN4957], to the case of college recruiting.

Geomarket search filters, the focus of this manuscript, slice states and metropolitan areas into smaller, local recruiting markets. For example, geomarket "CA08" refers to "Alameda Country excluding Oakland." A university might purchase prospect contact information by filtering on this geographic region in combination with additional filters, such as year of high school graduation and SAT test score. When we initiated this manuscript, we asked ourselves, "where did geomarket filters come from?" The answer surprised us. We had imagined ourselves studying the application of the "data science revolution" to college recruiting, but it turned out we were studying the 1970s and 1980s! The Market Segment Model and geomarkets can be read as the application of social science research methods to "capture and quantify the phenomenon behind the folklore" [@RN4982, p. 11] of college admissions officers. This work, comissioned by College Board, was subsequently turned into a market research tool -- The College Board Enrollment Planning Service -- that helps colleges decide where to look for prospective students.

@@

<!--Historically, College Board partnered with Members to develop projects that later became revenue-generating products for College Board.-->

@@

In 1978, Robert Zemsky, a University of Pennsylvania professor, was asked by the President to figure out, "'Who thinks about Penn?'" and "'What other institutions do they think about when they think about us?" [@RN4982, p. x]. To answer these questions, Zemsky began working with the Market Research Committee of the Consortium on Financing Higher Education (COFHE). As the project project became more ambitious, the researchers reached out to College Board because "we needed a database that described most institutions and most students\ldots Coincidentally, the Board was reviewing its own efforts to help colleges estimate their enrollment potential, efforts which had faltered largely because the smallest demographic unit used in tehse analyses was the state' [@RN4982, p. x]. In 1979, College Board began providing data and funding for what became the *Comprehensive Undergraduate Enrollment Planning Project (CUEPP)*.[^3] @RN4982 [p. 4] write that,

[^3]: say that Pennsylvania Department of Education was also a partner; see page x of Zemsky

@@

**Creating the Market Segment Model.** Considering the @RN5007 \[p. 402\] definition of quantification as "the production and communication of numbers," the Market Segment Model is an effort to quantify and communicate the knowledge that admissions officers have about student demand. @RN4982 [p. 4] state that "our initial task was to define enrollment markets in a manner consistent with admissions officers' intuitive understanding of student pools." They created "three types of boundaries" [@RN4982, p. 11], region, state, and community. The three regions were New England, Middle States, and the South. Next, "we divided each state into as few as two and as many as thirty community-based enrollment markets or pools, for a total of 143 separate markets" \[in New England, Middle States, and the South\].

These enrollment markets, later called "geomarkets" by the College Board Enrollment Planning Service were intended to be consistent with the conception of a catchment market from the perspective of admissions counselors. We know little about how geomarkets were created. @RN4982 write that,

@@

The two primary outputs of the market segment model are the (1) Market Segment Profile and (2) the Institutional Profile. The Market Segment Profile, which was produced separately for each of the 143 geo-markets, provides information about the number and characteristics of students in each market segment. @tbl-market-segment-characteristics reproduces a partial, simplified version of the bottom panel of @RN4982 [Figure 2.1] which represents the Market Segment Profile for "Connecticut Market 3: Fairfield County." @tbl-market-segment-characteristics has separate columns for each market segment (local, in-state, regional, national) and rows show the number of test-takers and characteristics of test-takers. For example, there were 550 "local" students in Fairfield county and these students submitted SAT scores to 2.8 postsecondary institutions on average. By contrast, there were 1,664 "regional" students who submitted test scores to 4.8 institutions on average. For local students, 11.2% had family income greater than \$35,000 and 9% had both parents with a BA. For regional students, 41.9% had family income greater than \$35,000 and 34.0% had both parents with a BA. Each Market Segment Profile also present information about the institutions that students from each segment sent scores to.[^4]

[^4]: @tbl-market-segment-institutions recreates the top panel of @RN4982 [Figure 2.1] for the Fairfield County, CT local market. For example, of the 550 local students, 277 (50.4%) sent scores to institution #1, which was a private master's granting institution. Of the 1,199 in-state students, 757 (63.1) sent scores to institution #1, a public doctoral granting institution, and 515 (43.0%) sent scores to institution #2, a public master's granting institution.

INSERT @tbl-market-segment-characteristics ABOUT HERE

INSERT @tbl-sample-institutional-profile ABOUT HERE

>>>>

---
title: "Structuring College Access: The Market Segment Model and College Board Geomarkets"
author: ["Ozan Jaquette (UCLA)", "Karina Salazar (University of Arizona)"]
bibliography: ./../assets/bib/cb_geomarket_bib.bib
csl: ./../assets/bib/apa.csl
format:
  pdf:
    documentclass: article
    #papersize: letter
    number-sections: true
    number-depth: 2
    #toc: FALSE
    #pandoc_args: !expr rmdfiltr::add_wordcount_filter(rmdfiltr::add_citeproc_filter(args = NULL))
    #keep_tex: true
    extensions: [andrewheiss/wordcount]
editor: visual
geometry:
  margin=1in
fontsize: 11pt
header-includes:
  - \usepackage{setspace}
  - \doublespacing
---

@@

In 1983, Robert Zemsky and Penney Oedel authored *The Structure of College Choice*. Based on an analysis of 1980 SAT score-sending behavior and published by *The College Board*, @RN4982 developed the Market Segment Model to help colleges recruit students. The underlying idea of the Market Segment Model is that student demand for your institution is fundamentally a function of class. Therefore, the key to recruitment is identifying territories where large numbers of your class live. @RN4982 [p. 11] describe the Market Segment Model as an effort to "quantify" the knowledge that college admissions officers have about the geography of student demand:

@@

Geomarkets have been incorporated into the supply-side structure of college access in ways that are simultaneusly mundane and fundamental. We posit three ways. First, Geomarkets became a principle for how many -- if -- admissions offices organized their labor. In states where a college recruits heavily from, admissions officers are often assigned specific Geomarkets as their territory \[CITE\]. Geomarkets were created to mirror the territories of admissions officers and, in turn, admissions offices structured their territories around Geomarkets. Second, The Market Segment Model and Geomarkets became the basis for Geomarkets are the basis for the College Board Enrollment Planning Service (EPS), founded in 1984 and still active today. EPS software recommends which Geomarkets a college should recruit from and which schools/communities they should prioritize within targeted Geomarkets. @RN5019_2010 says "Enrollment Planning Service (EPS\textsuperscript{\textregistered}) the analysis tool that pinpoints the schools and geomarkets where your best prospects are most likely to be found." @RN5021 reports that, in 1995, 37% of 4-year public institutions and 49% of 4-year private institutions used EPS.

@@

Geomarkets have never been the subject of empirical research. This omission is surprising because the sociology of education is concerned with sorting [@RN4873], particularly sorting into college. Traditionally, most scholarship -- and certainly most scholarship in economics \[HOXBY? CHETTY\] -- views inequality in college access as a function households or k-12 schools, which represent the demand-side of higher education. More recently, scholarship in the sociology of education takes seriously the idea that colleges are not passive recipients of applications, but rather they expend substantial resources soliciting demand from desirable applicants [@RN3519; @RN4774; @RN4324]. However, most scholarship on recruiting analyzes the behavior of individual colleges [e.g., @RN4758; @RN4759], thereby creating the assumption that recruiting is a function of individual colleges. What these studies have missed is the broader enrollment management industry -- third-party vendors, their products and consulting solutions -- as a set of mechanisms that structure college access, funneling certain kinds of students to certain kinds of institutions.

This manuscript analyzes one of those mechanisms, The Market Segment Model and College Board Geomarkets, as a case study of quantification. We develop arguments from the broader literature on quantification [e.g., @RN5003; @RN5005; @RN5007], particularly the discussions of correlation and homophily from @RN4975. Predictive analytics are based on correlation. When we use data on past correlations to make recommendations about the future, we amplify the effects of historic structural inequality. The Market Segment Model analyzes SAT score-sending data from 1980 and concludes that student demand for higher education is primarily a function of social class [@RN4982]. The Market Segment Model argues that homophily -- actors that share characteristics are likely to form connections -- is the organizing principle of student demand and competition between colleges. @RN4982 [p. 72] state that "the hierarchical structure of collegiate competition largely reflects the stratified social and economic dimensions of the communities from which colleges draw their students." This snapshot of student demand in 1980, itself a consequence historic structural inequality, was programmed into the way admissions offices organize labor and programmed into the EPS and Student Service Service products that colleges utilize to identify and target prospective students. The result is a supply-side that reinforces structural inequalities observed on the demand-side.

\[ADD TEXT SOMEWHERE AROUND HERE ABOUT WHY THIS TOPIC IS SIGNIFICANT AND IMPORTANT FOR CURRENT/FUTURE POLICY DEBATES: Why market segment model important. Software as service products dominate em industry and structure college access. Look at eab machine learning product But rarely do we see the recipe underneath these products. The book structuring college access is that recipe for EPS software\]

@@

```{=html}
<!-- 
THIS TEXT IS FINE, BUT TRYING TO SET UP THE CONTRIBUTION TO BROADER QUANTIFICATION LITERATURE, VIS-A-VIS NO-ONE HAS STUDIED MARKET RESEARCH; DON'T THINK THIS IS NECESSARY AND YOU ARE NOT ON FIRM GROUND HERE IN CALLING THIS "MARKET RESEARCH"
Scholarship on quantification in education has focused on consumer-facing measures such as U.S. News Rankings [@RN5003] and UK "league tables" about school quality [@RN5005]. To our knowledge, prior knowledge has not examined business-facing measures produced by third-party market research. Here, the desirable attributes of students and places they live are quantified with the purpose of informing colleges about where they should allocate recruiting resources to efficiently enroll the most desirable students. We reason that market research reproduces inequality by encouraging colleges to plan their recruiting efforts based on a snapshot of social stratification in student demand that is substantially a function of historic, structural inequality in educational opportunity. In contrast to @RN5005, we cannot demonstrate the effect of geomarkets on inequality. However, we show the mechanisms of how geomarkets reproduce inequality by simulating who is included/excluded when geomarkets are utilized within student list purchases as recommended by @RN4982.
-->
```

@@

A note from RESULTS that I texted to Bastedo on 12/19/2024

-   But these geomarkets are pretty fucked up. They were designed to capture class. I see lots of orders that condition on test scores and rich geomarkets. But I know from similar purchases that don’t condition on geomarket, that the high scoring first gen Asian and Hispanic students tend to come from the poor geomarkets that get excluded from a lot of purchases. By contrast, The Asian and Hispanic students that do get included in purchases that target rich geomarkets are usually not first gen

@@

Scholarship at the nexus of enrollment management and college access can be categorized by which part(s) of the enrollment funnel it speaks to. The majority of scholarship focuses on the admissions stage, analyzing which admissions criteria are utilized and/or which applicants are admitted [e.g., @RN4874; @RN4993; @RN4131; @RN3522; @poon2022rethinking]. Economists often investigate financial aid leveraging, which seeks to convert admits to enrolled students \[CITE\]. <!--  Scholarship on college access often takes a demand-side perspective, whereby household differences in economic and cultural capital yield stratification in student preferences and the ability to gain admission [e.g., @RN4935; @RN1814].   -->

@@

Scholarship about U.S. News & World Report (USNWR) Law School rankings demonstrate the reactivity, discipline, and authority effects of quantification [@RN5003; @RN2356; @RN2376; @RN2375]. Law school rankings were developed as a means of informing prospective students about the relative quality of different programs. Speaking to *reactivity* effects, rankings had the intended effect of affecting where students apply and enroll \[CITE\]. However, rankings also affected hiring decisions of firms because clients evaluated firms based on the prestige of law schools attended by firm lawyers. Once law schools realized that "important groups of constituents --- students, faculty, trustees, employers, other media --- were using rankings to make decisions that had large consequences for schools \ldots \[then\] schools felt pressured to take them seriously" [@RN5007, p. 415]. Rankings *disciplined* the behavior of law schools. For example, applicants with characteristics valued by the rankings system (e.g., LSAT scores) became more important for decisions about admissions and merit aid [@RN5003]. Rankings also weakened the *authority* of admissions directors and admissions officers to make independent decisions about which applicants to admit, ideal class size, and curricular offerinags [@RN5003]. <!--
In a process akin to sports analytics in Moneyball \[CITE MICHAEL LEWIS\], we suggest that market research replaces the wisdom and authority of local admissions counselors.    -->

@@

```{=html}
<!-- THIS IS THE CUT TRANSITION PARAGRAPH
In the broader quantification literature, critical data studies [e.g., @RN4975; @RN4849; @RN4772] and sociological analyses of technology and inequality [e.g., @RN4778; @RN4786; @RN4877; @RN4810] introduce concepts salient to the analysis of market research. Our focus is on "selection devices," which utilize some combination of inputs and discretion to categorize places, people, and organizations for the purpose of allocating opportunities or interventions [@RN4778; @RN5012].
-->
```

@@

Several studies show that using correlational models make predictions reproduces structural inequality [@RN4794]. The correlations observed during the training data stage are a snapshot of relationships between variables at a particular point of time. These correlations may be a function of enduring structural inequality, but underlying causes are not considered by applications of predictive models. Reviewing scholarship about algorithms, @RN4794 [p. 224] state that "predicting the future on the basis of the past threatens to reify and reproduce existing inequalities." Disproportionately targeted/excluded populations are predicted to have a higher risk of an outcome, which amplifies subsequent targeting/exclusion. This phenomenon whereby has been termed the "ratchet effect" [@RN4851] and "pernicious feedback loops" [@RN4849].[^2]

[^2]: An often cited example is the LSI-R recidivism model which predicts a prisoner's chances of re-arrest and is used by 24 states [@RN4849]. Because the algorithm uses zip code as an input, people who live in highly policed neighborhoods have a higher predicted probability of being arrested, which leads to more policing in those neighborhoods, which perpetuates racialized inequality in arrests. Note that predictive models such as the LSI-R model are not merely used for social science purposes. Rather, they reproduce structural inequality because they direct the allocation of future resources based on a snapshot of historical inequality.

**Homophily**. @RN4975 argues that homophily is central to models that use correlation to predict future behavior. Homophily means that actors who share common characteristics are likely to form connections with one another, or "birds of a feather flock together." Homophily is a core assumption for network science [@mcpherson2001birds], in which actors -- in this manuscript, high school students -- are linked to one another through direct and indirect network ties -- submitting SAT scores to the same college. Because network science models often draws from rational choice theory, they assume that homophily is the result of voluntary action by individuals, thereby "eras\[ing\] historical contingencies, institutional discrimination, and economic realities" [@RN4975, p. 95] that underlie behavior consistent with homophily. @RN4975 problematizes the idea that homophily is a naturally occurring phenomenon. In commercial social networks (e.g., Faceboo, X, TikTok), homophily is more than an assumption; homophily is programmed into the algorithms that create connections between users.[^3]

[^3]: Thus, @RN4975 [p. 82] writes, "echo chambers are not unfortunate errors, but deliberate goals" because "homophily is used to create agitated clusters of individuals whose angry similarity and overwhelming attrication to their common object of hatred both repel them from one another and glue them together."

Within the market research profession, homophily is central to geodemography and the creation of market segments. Geodemography emerged in the 1970s as a branch of market research that estimates the behavior of consumers based on where they live [@RN4878]. Market segments are subgroups within a larger market that have similar consumer demand. Early geodemographic classifications of consumers (e.g., PRIZM by Claritas Corporation) were derived from publicly available Census data, which disaggregated data to the zip code level. The Claritas Corporation had a financial incentive to argue that people living near one another share similar consumer preferences because geographic localities could then be categorized into market segments that would be useful for direct mail marketing campaigns [@RN4957]. Later, the development of individual credit scores (e.g., FICO score) enabled merchants to classify consumers into many, fine-grained groups [@RN4877]. @RN4810 introduce the concept "classification situations" to describe the expansion of actuarial techniques to categorize customers into many, ordinally ranked groups. Merchants and lenders began tying these classifications to tiered products that targets different consumer groups with different levels of benefits and costs [@RN5015].[^4] Classification situations engender markets where a vertical hierarchy of products are targeted to a vertical hierarchy of consumers. <!--
ORIGINAL TEXT ABOUT GEODEMOGRAPHIC CLASSIFICATION SYSTEMS

@@

[^4]: For example, "payday loans" charge high interest rates to consumer groups that were previously denied credit altogether.

Reviewing scholarship on quantification, @RN5006 argue that quantification has effects to the extent that stakeholders care about the numbers. U.S. law school rankings [@RN5003] and UK school league tables [@RN5005] exemplify this sort of salience. Studies of quantification in the sociology of education tend to study consumer-facing metrics. Law school rankings can be conceived as a business-to-consumer product because the goal is to inform consumers (prospective law school students) and consumers pay for the product. By contrast, market research helps businesses identify/target customers and can be conceived as a business-to-business (B2B) product because businesses pay for market research. To our knowledge, scholarship on quantification in education has not examined how third-party market research classifies consumers. Additionally, scholarship on college access has failed to analyze third-party products that colleges use to identify and target prospective students. This manuscript analyzes Market Segment Model [@RN4982], which categorized high school students into vertical market segments and simultaneously created local Geomarkets that could be evaluated based on their composition of student market segments. The Market Segment Model and Geomarkets became the basis for the College Board *Enrollment Planning Service* (EPS), which advised colleges which Geomarkets to target. Later, Geomarkets were incorporated into the College Board student list product named Student Search Service.

@@

```{=html}
<!--
The market segment model became the basis for the College Board *Enrollment Planning Service* (EPS), which advised colleges which Geomarkets to target. @RN5021 reports that in 1995, 37% of 4-year public universities and 49% of 4-year private universities used EPS, while 41% of 4-year publics and 16% of 4-year privates used ACT's market analysis service product, which was similar to EPS. Geomarkets later became a filter in the College Board *Student Search Service* student list product, which is utilized by the vast majority of public and private universities.

By contrast, many quantification projects -- for example, human rights indicators -- have "little oomph" [@RN5006, p. 265]. FROM BERMAN & HIRSCHAN PAGE 263: "Merry demonstrates how human trafficking indicators largely sit on the shelf because no audience for them has been created." 
-->
```

@@

```{=html}
<!--Historically, College Board partnered with Members to develop projects that later became revenue-generating products for College Board.

@@

The Market Segment Model was an effort to quantify and communicate the knowledge that admissions officers have about student demand. 
-->
```

**Creating the Market Segment Model.** In 1978, Robert Zemsky, a University of Pennsylvania professor, was asked by the President to figure out, "'Who thinks about Penn?'" and "'What other institutions do they think about when they think about us?" [@RN4982, p. x]. To answer these questions, Zemsky began working with the Market Research Committee of the Consortium on Financing Higher Education (COFHE), a consortium of 30 selective private universities founded in the mid-1970s. The researchers reached out to College Board because "we needed a database that described most institutions and most students\ldots Coincidentally, the Board was reviewing its own efforts to help colleges estimate their enrollment potential, efforts which had faltered largely because the smallest demographic unit used in these analyses was the state' [@RN4982, p. x]. In 1979, College Board began providing data and funding for what became the *Comprehensive Undergraduate Enrollment Planning Project (CUEPP)*. @RN4982 [p. 4] write that,

@@

@RN4982 \[p. 4\] state that "our initial task was to define enrollment markets in a manner consistent with admissions officers' intuitive understanding of student pools." Quoting an admissions officer, @RN4982 [p. 11] write, "'There are only three kinds of colege-bound students: those who want to live at home, those who want to live on campus but bring their laundry home, and those who want to go far enough from home that Mom and Dad can't visit without calling first.'" As such, [@RN4982] created "three types of boundaries" [@RN4982, p. 11], region, state, and community. The three regions were New England, Middle States, and the South. Next, "we divided each state into as few as two and as many as thirty community-based enrollment markets or pools, for a total of 143 separate markets."

These enrollment markets, later called "Geomarkets" by the College Board Enrollment Planning Service were intended to be consistent with the conception of a catchment market from the perspective of admissions counselors [@RN4982]. We know little about how geomarkets were created. @RN4982 write that,

@@

The two primary outputs of the market segment model are the (1) Market Segment Profile and (2) the Institutional Profile. Appendix A describes these outputs in more detail. The Market Segment Profile provides information about the number and characteristics of students in each market segment. This information is produced separately for each of the 143 Geomarkets. For example, Appendix @tbl-market-segment-characteristics shows how many students in "CT3 -- Fairfield County" are defined as regional or national based on their SAT score-sending behavior. The Institutional Profile shows the number of students who send scores to a particular institution -- separately by market segment -- and which majors these students are interested in. A university could obtain this for their own institution or for a competitor. Appendix @tbl-sample-institutional-profile reproduces a partial, simplified version of @RN4982 [Figure 2.3], the Institutional Profile of an anonymous college for students from Fairfield County, CT. This table shows that 69 Regional students and 109 National students from Fairfield County sent SAT scores to the college.

**Enrollment Planning Service**. In 1984, College Board created the Enrollment Planning Service (EPS) based on the Market Segment Model [@RN4987; @RN4980]. EPS was a simple software-as-service platform that informed college decisions about recruiting by essentially recreating the analyses of @RN4982. Institutions purchasing EPS could obtain the Market Segment Report for each local market and the Institutional Profile -- their own and that of competitors -- for each local market. Typical marketing material describes EPS as,

> The marketing software that pinpoints the schools and geomarkets where your best prospects are most likely to be found. With the click of a mouse, EPS provides you with comprehensive reports on your markets, your position in those markets, and your competition. Focus your valuable time and resources on the right prospects [@RN5019].

```{=html}
<!--
Promotional material highlights the themes of "locate the students you want" [@RN5019_2003] and "focus your valuable time and resources on the right prospects" [@RN5019]. @RN5019_2010 promotional material describes EPS as:

> the analysis tool that pinpoints the schools and geomarkets where your best prospects are most likely to be found. You can research your existing (or new) markets using selection criteria and locate your top prospects in various ways -- by states, geomarkets, counties, zip codes, high schools and international regions. EPS provides you with comprehensive reports on your markets, your position in those markets, and your competition."
